\message{ !name(../ImprovisationComputersInteraction.tex)}\makeatletter
\@ifundefined{Status}{\gdef\Status{0}}{}
\makeatletter
\makeatletter
\@ifundefined{ForPrinting}{\gdef\ForPrinting{0}}{}
\makeatletter
\ifcase\Status %{0}

%%% We are running tex4ht

\typeout{HTML output}
\documentclass[a4paper]{book}
%% Helvetica
\newcommand{\musicannot}{\fontsize{7}{9}\usefont{T1}{phv}{m}{n}}
\or % Status = 1

%%% We are running latex
\typeout{PDF output}
\documentclass[a4paper]{book}
% \renewcommand{\rmdefault}{pad}
% \renewcommand{\sfdefault}{pfr}
\newcommand{\musicannot}{\fontsize{7}{9}\usefont{T1}{phv}{m}{n}}
\fi

\usepackage[british]{babel}
% \usepackage[T1]{fontenc}

\usepackage[babel]{csquotes}
\usepackage{epigraph}
\setlength{\epigraphwidth}{8cm}

%%%%% Biliography style %%%%%
%style=verbose-ibid,
\usepackage[style=verbose-ibid,
            indexing=cite,
            sorting=nyt]{biblatex}

\renewcommand*{\mkibid}{\emph}
% \renewbibmacro*{citeindex}{\indexnames{author}} 

\bibliography{bibliography}
\DeclareBibliographyCategory{nobib}
\addtocategory{nobib}{cybernetic08,concours-spat,wiki_synklavier,j2ee,j2se}
% \DefineBibliographyStrings{swedish}{}
% \DefineBibliographyExtras{british}{}

%,bateson72,bateson72:cyber,bateson72:cyber-self,bateson72:play

\usepackage[usenames]{color}
%\usepackage[pdftex]{graphicx}
\usepackage[a4paper]{geometry}
\usepackage{fancyhdr}
\usepackage{setspace}
\onehalfspacing
%\usepackage[final]{pdfpages}
\usepackage[toc,page]{appendix}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{rotfloat}
\usepackage{enumerate}
\usepackage{sectsty}
\usepackage{makeidx}
\makeindex

\usepackage[hypertoc=true,hyper=true]{glossary}
\makeglossary

%%% etherSound %%%
\usepackage{amsmath}
\numberwithin{equation}{section}
\usepackage{lscape}
\usepackage{tabularx}
%%% /etherSound %%%

\definecolor{svart}{rgb}{0,0,0}
\usepackage[
     draft=false,
     a4paper=true,
     plainpages=false,
     colorlinks=true,
     citecolor=svart,
     linkcolor=svart,
     urlcolor=svart,
     pdftex,
     pdfpagelabels]{hyperref}
\urlstyle{same}

\usepackage[all]{hypcap}

\fancyhead{}
\ifcase\ForPrinting%{0}
\typeout{Header links}
\fancyhead[LO]{\hyperref[sec:toc]{{\small Contents}}}
\fancyhead[RE]{\hyperref[sec:biblio]{{\small Bibliography}}, \hyperref[sec:index]{{\small Index}}}
\or
\typeout{No header links}
\fi
\fancyhead[RO]{\nouppercase{\small{\textit{\leftmark}}}}
\fancyhead[LE]{\nouppercase{\small{\textit{\rightmark}}}}

\hyphenation{op-tical net-works semi-conduc-tor collabo-rations Musik-sociolo-gin Mu-sik-soziolo-gie super-re-ella}

%%% Different font in captions
\newcommand{\captionfonts}{\footnotesize}
\makeatletter  % Allow the use of @ in command names
\long\def\@makecaption#1#2{
  \vskip\abovecaptionskip
  \sbox\@tempboxa{{\captionfonts #1: #2}}
  \ifdim \wd\@tempboxa >\hsize
    {\captionfonts #1: #2\par}
  \else
    \hbox to\hsize{\hfil\box\@tempboxa\hfil}
  \fi
  \vskip\belowcaptionskip}
\makeatother   % Cancel the effect of \makeatletter

\renewcommand{\figurename}{Fig.}

\newcommand{\listoftags}{\fontsize{7}{9}\usefont{T1}{phv}{m}{n}}
\newcommand{\tagword}[2][]{#2}

%%% Use tagwords like this: %%%
%%% {\listoftags \tagword{These} \tagword{Are} \tagword{Tag} \tagword{Words}}

%%% Font styles %%%

\makeatletter

\def\section{\@startsection {section}{1}{\z@}{-3.5ex plus -1ex minus -.2ex}{2.3ex plus .2ex}{\usefont{T1}{phv}{m}{n}}}
\def\subsection{\@startsection{subsection}{2}{\z@}{-3.25ex plus -1ex minus -.2ex}{1.5ex plus .2ex}{\usefont{T1}{phv}{m}{n}}}
\def\subsubsection{\@startsection{subsubsection}{3}{\z@}{-3.25ex plus -1ex minus -.2ex}{1.5ex plus .2ex}{\usefont{T1}{phv}{m}{it}}}
\makeatother

\chapterfont{\usefont{T1}{phv}{m}{n}}

\renewcommand{\appendixpagename}{\usefont{T1}{phv}{m}{n} Appendices: Published papers and documentation of artistic work.}
\newcommand{\HRule}{\rule{\linewidth}{0.3mm}}

%%% quote environment %%%

\newenvironment{squote}{
\begin{quote}\small}{
\end{quote}}

\newcommand\T{\rule{0pt}{2.6ex}}
\newcommand\TT{\rule{0pt}{3.6ex}}
\newcommand\B{\rule[-1.2ex]{0pt}{0pt}}
\newcommand\BB{\rule[-2.2ex]{0pt}{0pt}}

\input{include/ImpComInt-glossary.tex}

\title{\usefont{T1}{phv}{m}{n}Improvisation, Computers, and Interaction\\{\small Rethinking Human-Computer Interaction Through Music}}
\author{Henrik Frisk}
%\date{\today}
\date{}
\begin{document}

\message{ !name(include/ImpComInt-abstract.tex) !offset(-157) }
\chapter*{Abstract}
\label{cha:abstract}
\addcontentsline{toc}{chapter}{Abstract}
\setcounter{page}{1}
\pagestyle{plain}

Interaction is an integral part of all music. Interaction is part of listening, of playing,
of composing and even of thinking about music. In this thesis the multiplicity of modes in
which one may engage interactively in, through and with music is the starting-point for
rethinking \index{interaction!human-computer}\index{HCI}Human-Computer Interaction in general
and Interactive Music in particular. I propose that in
\index{interaction!human-computer}\index{HCI}Human-Computer interaction the methodology of
control, \emph{\index{interaction!as control}interaction-as-control}, in certain cases should
be given up in favour of a more dynamic and reciprocal mode of interaction,
\emph{\index{interaction!as
difference}\index{interaction-as-difference}interaction-as-difference}: Interaction as an
activity concerned with inducing differences \emph{that make a difference}.
\emph{\index{interaction!as
difference}\index{interaction-as-difference}Interaction-as-difference} suggests a kind of
parallelism rather than click-and-response. In essence, the movement from \emph{control} to
\emph{difference} is a result of rediscovering the power of improvisation as a method for
organising and constructing musical content. \emph{Control} and \emph{difference} are not
mutually exclusive, however; they are not opposing concepts. \emph{Interaction-as-difference}
is to be understood as a broadening of the more common paradigm of direct manipulation in
\index{interaction!human-computer}\index{HCI}human-computer interaction. Improvisation is at
the heart of all the sub-projects included in this thesis, and also, in fact, those that are
not immediately related to music but geared more towards computation. Trusting the
self-organising aspect of musical improvisation, and allowing it to diffuse into other areas
of my practice, constitutes the pivotal change that has radically influenced my artistic
practice. The \useGlosentry{glos:work_in_movement}{\emph{work-in-movement}}, a concept
introduced by Umberto Eco, is (re-)introduced as a work type that encompasses radically open
works. The \emph{work-in-movement}, presented and exemplified by a piece for guitar and
computer, requires different modes of representation, as the traditional musical score is too
restrictive and is not able to communicate that which is the most central aspect: the
collaboration, negotiation and interaction. The Integra framework---comprising a database
model and a corresponding XML representation---is proposed as a means to produce annotated
scores that carry past performances and versions with it. The notion of the giving up of the
\index{Self}Self is suggested as the common nominator, the prerequisite, for an improvisatory
and self-organising attitude towards musical practice that allows for
\emph{\index{interaction!as
difference}\index{interaction-as-difference}interaction-as-difference}. Only if the
\index{Self}Self is able and willing to accept the loss of priority of interpretation,
willing to give up or disregard faithfulness to ideology or idiomatics, is difference
conceivable. Only if one is willing to \emph{forget} is \emph{\index{interaction!as
difference}\index{interaction-as-difference}interaction-as-difference} made possible. Among
the artistic works that have been produced as part of this inquiry are some experimental
tools in the form of computer \index{software}software to support the proposed concepts of
interactivity and together with the musical works they make up both the object and the method
in this PhD project. Contained within the thesis, these sub-projects (all of which are
works-in-progress), are used to make inquiries into the larger question of the significance
of interaction in the context of artistic practice involving computers.

\message{ !name(../ImprovisationComputersInteraction.tex) !offset(91) }

\end{document}