\storeglosentry{glos:sociotech}
{name={Sociotechnical systems},
  description={Sociotechnical systems theory is theory about the social aspects of people and society and technical aspects of machines and technology.} Sociotechnical systems}

\storeglosentry{glos:esthesic}
{name={Esthesic},
  description={In musical semiology, according to Nattiez, the phase of interpretation or reconstruction of a musical work.}}

\storeglosentry{glos:icmc}
{name={ICMC},
  description={International Computer Music Conference. A yearly conference dedicated to computer music. Organized by ICMA, International Computer Music Association.}}

\storeglosentry{glos:poietic}
{name={Poietic},
  description={According to musical semiologists Jean-Jacques
  Nattiez and Jean Molino, the poietic phase of a musical work is the
  stage at which the musical material is constructed. According to
  \citeauthor{nattiez}, articulating the poietic and esthesic level
  ``facilitates knowledge of all processes unleashed by the musical
  work''.\footcite[92]{nattiez} We reinterpret the terms `poietic'
  and `esthesic' in the \emph{Negotiating the Musical Work} papers. (see \cite{frisk-ost06} and \cite{frisk-ost06-2})}}

\storeglosentry{glos:musician}
{name={Musician},
  description={I use `musician' in a very inclusive
  way in these texts. A composer, an improviser, a performer are all
  sub-categories to the general description `musician'. I am leaning towards Trevor Wishart who, in his book \citetitle{wis96}, points to how the increased importance given to notation has split the musician in two: the composer and the performer. In my own practice it is a goal in itself to unite these two roles.}}

\storeglosentry{glos:six-tones}
{name={The Six Tones},
  description={A musical collaboration between myself, Stefan \"{O}stersj\"{o}, Ngo Tra My and Ngyen Thanh Thuy started in 2006. The group was named after the piece that was collaboratively constructed and entitled \emph{The Six Tones}}}

\storeglosentry{glos:MIDI}
{name={MIDI},
  description={A low band-width communication protocol enabling electronic musical devices (instruments, controllers, computers, etc.) to communicate with one another. MIDI was not initially designed for the purposes it is now used. OSC is a similar but more contemporary and flexible protocol.}}

\storeglosentry{glos:OSC}
{name={OSC},
  description={Open Sound Control. A communication protocol for connecting electro-acoustic music devices such as computers, sound synthesizers and other multi media devices over TCP/IP primarily using UDP. A sort of successor to MIDI.}}

\storeglosentry{glos:csound}
{name={Csound},
  description={A programming language designed for signal processing. Primarily intended for use in non-real time.}}

\storeglosentry{glos:electro-mus}
{name={Electro-acoustic Music},
 description={Also EAM. According to the Oxford English Dictionary electro-acoustics are ``acoustics investigated by electrical methods'' (``electro-acoustics'' The Oxford English Dictionary. 2nd ed. 1989. OED Online. Oxford University Press. 31 Oct. 2007). Electro-acoustic music is a broad term used to denote music produced  by or with electrical methods. Today, since this is mainly achieved  by the use of digital computers, in USA the use of the term `Computer Music' is more common. Yet another term, more commonly used in the francophone countries is `acousmatic', and it is sometimes argued that `acousmatic' refers to the genre and `electro-acoustic' to the means of production.\footcite[E.g.][\textparagraph~9]{musique-recherche} Throughout this text I am primarily using electro-acoustic music or electro-acoustics to denote my own artistic work involving computers and music.}}

%\footcites[See][]{EMS06}[in particular][]{landy06}{dack06}{battier06}[and in relation to translation see][]{fields06}. 
%()
\storeglosentry{glos:fixed_media}
{name={Fixed media music},
  description={Music that is pre-recorded and stored onto a fixed media of some sort (a tape, a CD, a DVD, a hard-drive). Was earlier, prior to CDs and the digital media revolution, commonly referred to as `tape music' since, at the time, the media both for production and for storage was a tape. On of the least interactive types of electro-acoustic music.}}

\storeglosentry{glos:latency}
{name={Latency},
  description={The response time of an interactive system. The time between casue and action, between stimulus and response.}}

\storeglosentry{glos:pitch-track}
{name={Pitch-tracking},
  description={To let a computer (or a special purpose device) analyze an audio signal in real-time and extract the most likely fundamental of the sound (also pitch-to-MIDI, pitch detection).}}

\storeglosentry{glos:agent}
{name={Intelligent agent},
  description={An idea to improve HCI introduced in the mid 90's. An agent is a piece of software designed to collect and sort information and present it to the user. The idea is that the agent will learn what it is its user wants, or needs to know about. (The `intelligent agent' should not be confused with the `agent' as a factor in the production of musical content.)}}

\storeglosentry{glos:Max}
{name={Max/MSP},
  description={Named after computer music guru Max Mathews, Max/MSP is a software originally developed by Miller Puckette at Ircam in Paris. It is a tool to support the design and implementation of interactive music systems and allow for programming using a graphical environment by creating and interconnecting boxes that encapsulate a specific function. MSP is the signal processing part of Max.}}

\storeglosentry{glos:pd}
{name={Pure Data},
  description={Just like Max/MSP, Pure Data is developed by Miller Puckette. It is a graphical programming environment to build `patches' of audio and control rate processes. Pd, unlike Max/MSP is Open source.)}}

\storeglosentry{glos:score_follow}
{name={Score following},
  description={Relates to the technique that allows a computer to `listen to' and `follow' a score in real-time in a way that resembles how human performers or listeners are able to recognize and follow a piece of music they know, even if the performer(s) temporarily deviates from the score. See \citeauthor{dannenberg89} (\citeyear{dannenberg89}), \citeauthor{dannenberg06} (\citeyear{dannenberg06}) and \citeauthor{rowe} (\citeyear{rowe}), chap. 3.2.}}

\storeglosentry{glos:work_in_movement}
{name={Work-in-movement},
  description={In his book \citetitle{eco68} (\emph{Opera Aperta}) Umberto \citeauthor{eco68} coined the term the ``work in movement'' as a special sub-category of the open work. It is used here to name the work type that (contrary to the predominantly modernist examples chosen by \citeauthor{eco68}) perpetually keeps moving, possibly in multiple directions simultaneously and preferably, but not necessarily, in a trajectory \emph{away} from the originator. The movement is part of its authenticity. See also \emph{interaction-as-difference}.}}

\storeglosentry{glos:dsp}
{name={DSP},
  description={Digital Signal Processing. The processing of a signal in the digital domain. In the context of music DSP usually refers to audio processing (filtering, modulation, echo, etc.) and synthesis (generation of new sound material) in a computer, synthesizer or dedicated hardware.}}

\storeglosentry{glos:som}
{name={SOM},
  description={Self-Organizing Map. A special type of ANN that adopts itself according to some feature of the input and outputs a topography of this feature. A SOM is used in the subproject \emph{timbreMap}}}

\storeglosentry{glos:ann}
{name={ANN},
  description={Artificial Neural Network. A simplified model of a ``network of neurons that occur naturally in the animal brain.'' \parencite[6]{gurney97} There are many different types of ANN and they may all be gathered under the umberella of \emph{connectionism}. See also SOM.}}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../ImprovisationComputersInteraction"
%%% End: 
