
\section{Introduction}
\label{sec:introduction}

When we engaged in the actual work of producing a piece of music for
10-stringed guitar and electronics we departed from the point of view
that the activities of either one of us could be analyzed and
understood as being poietic (constructive) or esthesic
(interpretative) and that both of us could be oscillating back and
forth between these two modes. (This oscillation was something we had observed in the
collaboration between Stefan {\"O}stersj{\"o} and composer Love Mangs
described in \citet{frisk-ost06}.) I.e. already at the outset had we
consciously dismantled part of the notion of the composer-performer
dichotomy with the main purpose of engaging in a
collaboration in which we would allow ourselves to challenge our
respective roles. Challenge how these roles affects our choices but
also how they influence the collaboration. As we will see this
attitude resulted in some unexpected results as well as (obviously)
some expected.

In a previous article \citep{frisk-ost06-2} we outlined a theorethical
background for an altered use of the poietic-esthesic dynamics and a
semiotic approach to analyzing communication between agents involved
in the production of musical content. In Section
\ref{sec:coll-work:-backgr} we will look at other aspects of
collaborations and collective work. In Section
\ref{sec:split-musician} we look further into the consequences of
reclaiming \emph{the musician} at the expense of the specialized roles
of the \emph{the composer}, \emph{the performer}, \emph{the
  programmer}, etc. The unfolding of the different versions of the
composition \emph{Repetition Repeats all other Repetitions} that we
hitherto have produced and that we plan to produce is described in
Section \ref{sec:instances}. Finally, in Section
\ref{sec:where-did-work} we discuss the elusive concept of ``the
work'' in a collaboration such as this and make an attempt at mapping
what lies in the future for this particular composition, but also for
other similar collaborations.

\section{Collective work: A background}
\label{sec:coll-work:-backgr}

\section{The Split of The Musician}
\label{sec:split-musician}

\section{The Instances}
\label{sec:instances}

\subsection{The Score}
\label{sec:score}

\subsection{The First Version}
\label{sec:first-version}

\subsection{The Second Version: Breaking the rule}
\label{sec:second-vers-break}

\subsection{The Third Version}
\label{sec:third-version}

\subsection{Rethinking the "work-in-progress"}
\label{sec:reth-work-progr}

\section{Where did the work go?}
\label{sec:where-did-work}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../Repetition"
%%% End: 
