
%\section*{\textsf{3. Research experience}}

\subsubsection*{\textsf{Research environment}}

I am or have been involved in the following research environments and networks:

\begin{enumerate}
\item \textbf{CATS}: An initiative to bring together KMH, KKH and Stockholm University to initiate and stimulate cross discipline projects. 
\item \textbf{IAC}: Inter Arts Center at Lund University is a network as well as a location for artistic research. It has been the home of several of my projects since it opened a few years ago.
\item \textbf{Integra}: Integra was a research project financed by the EU that I worked intensively with during the years it was active. Not only was it an active environment at the time but it also led to sustained contact with Birmingham, Oslo and McGill, Montreal, Canada.
\item \textbf{(Re)Thinking Improvisation}: A research project and a research environment active between 2008 through 2013.
\item \textbf{ICMC}: Up until 2013 I was quite active in ICMC and the conferences it produced and several people in ICMC are still part of my network.
\item \textbf{EPARM (AEC)}: This loosely built network, organized around a small working group of which I am a part, has proven very useful to my own projects in the way it gives me insight in the state of affairs concerning artistic research in Europe in general.
\end{enumerate}

\subsubsection*{\textsf{Supervision experience}}

\subsubsection*{\textsf{As main supervisor}}
  \begin{enumerate}[label=\arabic*]
  \item Anna Einarsson

    \begin{tabular}[c]{p{3.3cm} p{7cm}}
      \textbf{Exam}: & 2016 \\
      \textbf{School}: & KMH \\
      \textbf{Project}: & \emph{A study in voice and composition. New means of performance and computer assisted composition by singing voice feature recognition} \\
      \textbf{Secondary supervisor}: & Susan Kozel \\
    \end{tabular}

  \item Kent Olofsson

    \begin{tabular}[c]{p{3.3cm} p{7cm}}
      \textbf{Exam}: & 2016 \\
      \textbf{School}: & KMH \\
      \textbf{Project}: & \emph{Rethinking Music Drama: Composing Sonic Art Theatre} \\
      \textbf{Secondary supervisor}: & Erik Rynell \\
    \end{tabular}

  \end{enumerate}
\subsubsection*{\textsf{As secondary supervisor}}

  \begin{enumerate}[start=3]
  \item Susanne Rosenberg

    \begin{tabular}[c]{p{3.3cm} p{7cm}}
      \textbf{Exam}: & 2014 \\
      \textbf{School}: & KMH \\
      \textbf{Project}: & \emph{Kurbits-ReBoot, svensk folksång i ny scenisk gestaltning} \\
      \textbf{Main supervisor}: & Hannu Tolvanen \\
    \end{tabular}
  \item Sara Wilén

    \begin{tabular}[c]{p{3.3cm} p{7cm}}
      \textbf{Exam}: & 2016 \\
      \textbf{School}: & KMH \\
      \textbf{Project}: & \emph{Opera improvisation} \\
      \textbf{Main supervisor}: & Karin Johansson \\
    \end{tabular}

  \end{enumerate}


\subsubsection*{\textsf{Conferences and symposia organized}}

\begin{enumerate}
\item \textbf{EPARM Stockholm}

  \begin{tabular}[l]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Working group member & AEC & April 2014    
  \end{tabular}

\item \textbf{SMAC/SMC}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Music Co-chair & KMH/KTH & August 2013    
  \end{tabular}

\item \textbf{Att utveckla och kommunicera musikalisk kunskap }

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Co-chair & Kungl. Mus. Akademien & Februari 2013    
  \end{tabular}

\item \textbf{EPARM Lyon}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Working group member & AEC & April 2013    
  \end{tabular}

\item \textbf{KMH Music research days}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Co-chair & KMH & January 2013    
  \end{tabular}

\item \textbf{EPARM Rome}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Working group member & AEC & April 2012    
  \end{tabular}

\item \textbf{(Re)Thinking Improvisation: International sessions of artistic research}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Co-chair & MHM & November 2011    
  \end{tabular}

\item \textbf{Eye of the Needle}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Co-chair & MHM & October 2008
  \end{tabular}
\item \textbf{Connect}

  \begin{tabular}[c]{p{0.33\linewidth} p{0.33\linewidth} p{0.33\linewidth}}
    Associate & MHM & 2006    
  \end{tabular}

\end{enumerate}

\subsubsection*{\textsf{Editor experience}}

\emph{(Re)thinking Improvisation: Artistic explorations and conceptual writing} \citep{frisk-re_improv13}: A two hundred page book, two audio CDs and one DVD.

\subsubsection*{\textsf{Research collaborations}}

\subsubsection*{\textsf{Integra}}

Integra brought together research centers and new music ensembles and was led by Birmingham Conservatoire at Birmingham City University in the UK. The activities of the project included developing new software to make music with live electronics, modernizing works that use old technology, commissioning composers and overseeing performances. The project was funded by the EU started in 2006 and went on for six years. 

My main duties in the project was working with software development, participating in research meetings and collaborating with the other partners of the project.

\subsubsection*{\textsf{(Re)Thinking Improvisation}}

(Re)Thinking Improvisation was a research project started by the Malmö Academy of Music and financed by the Swedish Research Council and involved artists and researchers from Sweden, the Netherlands and Asia. The project started in 2009 and ended 2013. 

My role in the project was to coordinate it together with my colleague Stefan Östersjö. 

\subsubsection*{\textsf{Review experience}}

\subsubsection*{\textsf{As opponent}}
%% p{0.23\linewidth} p{0.23\linewidth} p{0.23\linewidth} p{0.23\linewidth}
\begin{tabular}[c]{llll}
  Kim Hedås & Dissertation & September 2013 & Göteborg University \\
  Peter Spissky & 25\% seminar & February 2012 & Malmö Academy of Music \\
  Sten Sandell & 25\% seminar & February 2010 & Göteborg University \\
\end{tabular}

\subsubsection*{\textsf{As reviewer}}

\begin{itemize}
\item KU - artistic development projects in Gothenburg 2011
\item Review of the publication \emph{Art Monitor} for the Artistic Faculty, Göteborgs Universitet.
\item Music submissions for SMC 2013, KMH/KTH
\item Paper sumbissions for EMS 2012, KMH
\item Peer reviews for JAR publications
\item Reviews for Critical Studies in Improvisation publications
\item Reviews for NIME
\item Reviews for ICMC
\item Reviews for (Re)Thinking Improvisation
\item Reviews of project proposals for Framtidens Kultur 2006-2010
\item Reviews of artistic development projects at Malmö Academy of Music, 2002-2013.

\end{itemize}

\subsubsection*{\textsf{Prices and rewards}}

No particular scientific prices.

\newpage

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "vetenskaplig-meritportfolj"
%%% End: 
