Enligt Lunds Univsersitets mall för Vetenskaplig meritportfölj finns inte konstnärliga arbeten med. För att göra denna uppställning tydlig, och så att den ändå ska följa mallen, har jag lagt de textbaserade forskningsarbeten jag gjort först enligt den efterfrågade uppställningen (4.1) och därefter listat de konstnärliga och de konstnärliga forskningsarbetena (4.2). Detta trots att de senare av hävd är själva grunden för de tidigare och att uppställningen därför borde vara i omvänd ordning.

\subsubsection*{\textsf{a) Publicerade originalartiklar i referee-bedömda internationella tidskrifter}}


\begin{enumerate}
\item \fullcite{frisk-ost13}

\item  \fullcite{frisk-cobussen09}

\item  \fullcite{frisk09:improv}

\item  \fullcite{frisk07}

\item  \fullcite{frisk05}
\end{enumerate}


\subsubsection*{\textsf{b) Översiktsartiklar och andra inviterade artiklar i internationella tidskrifter}}
\subsubsection*{\textsf{c) Böcker, bokkapitel}}


\begin{enumerate}
\item \fullcite{frisk12-improv}

\item   \fullcite{frisk-re_improv13}

\item   \fullcite{frisk2013}

\item   \fullcite{friskostersjo2013}

\item   \fullcite{friskostersjo2013b}

\item   \fullcite{frisk11:ELIA}

 \item  \fullcite{frisk10}
\end{enumerate}


\subsubsection*{\textsf{d) Övriga artiklar och rapporter publicerade i internationella tidskrifter}}
\subsubsection*{\textsf{e) Vetenskapliga artiklar och rapporter publicerade på svenska}}

\begin{enumerate}
\item \fullcite{frisk13-vr}
\end{enumerate}

\subsubsection*{\textsf{f) Populärvetenskapliga artiklar / presentationer}}


\begin{enumerate}
\item \fullcite{frisk14-ab}

\item  \fullcite{frisk2013-2}

 \item \fullcite{frisk2013c}

 \item \fullcite{frisk-better}

 \item \fullcite{frisk11}

 \item \fullcite{friskDiary}
\end{enumerate}


\subsubsection*{\textsf{g) Konferensbidrag }}


\begin{enumerate}
\item \fullcite{frisK-bull11}

\item  \fullcite{frisk09}

\item  \fullcite{frisk-bullock08}

\item  \fullcite{frisk-bull07}

\item  \fullcite{frisk-ost06}

\item  \fullcite{frisk-ost06-2}

\item  \fullcite{frisk1}
\end{enumerate}



\subsubsection*{\textsf{h) Manuskript (inskickade manuskript ska listas först följda av de under bearbetning)}}

\begin{enumerate}
\item \fullcite{frisk2015} (Under bearbetning, utgivning 2015)
\end{enumerate}

\newpage
\subsection*{\textsf{iii. Publikationer - konstnärliga}}

\subsubsection*{\textsf{Fonogram - urval}}

\begin{enumerate}
\item [2014] lim, \emph{No title}, Kopasetic (under production)
\item [2013] The Six Tones, \emph{Signal in Noise}, dB
  Productions (double CD)
\item [2013] (Re)Thinking Improvisation, \emph{Artistic explorations},
  Lund University Press (double CD)
\item [2011] Stefan Ostersjo, \emph{Strandlines}, dB Productions
\item [2011] lim, \emph{lim with Marc Ducret}, Kopasetic Prod.
\item [2011] Scandinavian Electroacoustic Music, \emph{Spanning},
  ChamberSound
\item [2008] Henrik Frisk \& Peter Nilsson, \emph{etherSound},
  Kopasetic Prod.
\item [2007] lim and Marc Ducret, \emph{KOPAlectric}, Kopasetic Prod.
\item[2006] lim, \emph{SuperLim}, Kopasetic Prod.
\item[2006] New Century Series, \emph{Volume 17}, MMC.
\item[2005] David Liebman Big Band, \emph{Beyond the line}, OmniTone
\item[2004] Viola con Forza, \emph{Henrik Frendin}, Phono Suecia.
\item [2003] The Orchestra plays Jakob Riis, \emph{In Absence of
    Mind}, Cope Records
\item[2003] lim, \emph{lim}, dB Productions
\item[2002] Fixerad Anarki, \emph{Fixerad Anarki}, dB Productions
\item [1999] NoBass, \emph{Hello World}, LJ Records
\item[1999] Expressions, \emph{Henrik Frisk/Richie Beirach},
  Hornblower Recordings
\item [1998] The Orchestra, \emph{Noxx}, dacapo
\item[1998] The Orchestra, \emph{SmokeOut}, dB Productions
\item[1997] Blue Pages, \emph{Blue Pages}, Caprice Records
\item [1996] Beijbom/Kroner Big Band, \emph{Live in Copenhagen}, Four
  Leaf Clover
\item[1995] Henrik Frisk, \emph{Inventions of Solitude}, Hornblower
  Recordings
\item [1993] The Orchestra, \emph{Not as softly as\ldots}, MusicMecca
\end{enumerate}


\subsubsection*{\textsf{Beställningar (urval)}}
\begin{enumerate}
\item [2013] Ars Nova: work for video and electronic music: \emph{Orpheus in motion}
\item [2011] G\"{o}teborgs Biennalen: music for Better Life by Isaac Julien: \emph{Better Life}
\item [2011] Ensemble Midt Vest: interactive production for chamber group, interactive electronics and video.
\item[2011] Kopasetic, Konstn\"arsn\"amnden: work for jazz ensemble and electronics: \emph{M\r{a}nens gr\r{a}a \"{o}gon}
\item [2010] The Trembling Aeroplanes: work for digital chamber group: \emph{The Mystic Writing Pad}
\item [2009] Ensemble Ars Nova, Swedish Arts Council: work for chamber group and electronics: \emph{The Transparent I}
\item[2008] Marcel Cobussen, music for a web site: \emph{The Field of Musical Improvisation}.
 \item[2006] Copenhagen Art Ensemble, small big band and computer: \emph{Continuous Breach}.
 \item[2005] Stefan \"Ostersj\"o for \emph{Repetition Repeats all other
 Repetitions}, 10-stringed guitar and computer.
 \item[2005] Mercedes Gomez (\emph{work in progress}), harp and
 computer.
 \item[2003] Miya Yoshida for \emph{etherSound}, mobile phone sound
 installation.
 \item[2001] Stockholm Saxophone Quartet for \emph{Perspicio}, Saxophone
 quartet and computer.
 \item[2002] Henrik Frendin for \emph{Drive}, Electric viola grande and
 computer. 
\end{enumerate}

\subsubsection*{\textsf{Konserter (urval)}}

\begin{enumerate}
\item [2014] \emph{Saxophone and computer} Concerts in California.
\item [2014] \emph{Saxophone} Concerts with lim in Sweden and Denmark.
\item [2013] \emph{Saxophone and computer} Hanoi Sound stuff and Tour in Vietnam.
\item [2013] \emph{Saxophone} Lennart \r{A}berg Group featuring Peter Erskine, UKK, Uppsala.
\item [2012] \emph{Soxophone and computer} Frisk/Frendin, Musik i Halland.
\item [2012] \emph{Soxophone} Stroman/J\"{o}nsson group, London, UK, and Sweden.
\item [2012] \emph{Saxophone} KOPAorchestra on tour, Sweden and Denmark
\item [2011] \emph{Saxophone and computer} International session on artistic research, Malm\"o.
\item [2011] \emph{Saxophone} KOPAorchestra on tour, Sweden
\item [2011] \emph{Saxophone} Stroman/J\"onsson Vocal project feat. Lena Willemark, KOPAfestival, Lund, Sweden
\item [2011] \emph{Saxophone and computer} Collaboration with Isaac Julien, G\"oteborg International Biennial for Contemporary Art, Sweden
\item [2011] \emph{Saxophone and computer} Kalvfestivalen, Kalv, Sweden
\item [2011] \emph{Computer} Summerfestival, Ensebmle Midt Vest, Herning, Denmark
\item [2011] \emph{Computer} Integra Festival, Copenhagen, Denmark
\item [2011] \emph{Saxophone and computer} re:New festival, Copenhagen
\item [2010] \emph{Composition} Repetition Repeats\ldots in Ghent, Bruxelles
\item [2010] \emph{Saxophone} Frisk / Beirach / Mogensen trio, tour in Sweden and Denmark.
\item [2010] \emph{Saxophone and computer} EarZoom festival, Lubljana, Slovenia.
\item [2010] \emph{Saxophone and computer} The Six Tones concert at Green Space Festival, Hanoi, Vietnam.
\item [2009] \emph{Computer} - Performance of Anne LeBaron's cyberopera \emph{Sucktion}
\item [2009] \emph{Saxophone and bandleader} - Premiere of the 10-piece band Henrik Frisk's Pli at Kopasetic Festival, Malm\"o, Sweden.
\item [2009] \emph{Computer} - Performances of \emph{Continuous Breach} at Copenhagen and Aarhus jazz festivals.
\item [2009] \emph{Computer} - etherSound in Stockholm at Nybrokajen 11.
\item [2009] \emph{Saxophone} - performances in Malm\"o and Lund with lim.
\item [2009] \emph{Computer} - Tour in Sweden and Denmark with The Six Tones
\item [2007] \emph{Saxophone} - Performances with lim and Marc Ducret. 
\item [2007] \emph{Saxophone and computer} - ICMC Copenhagen
 \item[Since 2006] \emph{Computer} - Performances with Stefan \"Ostersj\"o (guitar) in Vietnam, China, USA, UK and Sweden.
 \item[Since 2003] \emph{Saxophone} - Performances in Sweden and USA in duo with Per-Anders
 Nilsson (laptop).
 \item[Since 2002] \emph{Computer} - Performances in Sweden, Iceland,
 USA, Canada and Germany in duo with Henrik Frendin (viola).
 \item[2002] \emph{Saxophone} - Tour in Mexico.
 \item[2002] \emph{Saxophone and computer} - ICMC, Gothenburg, Sweden
 \item[2001] \emph{Saxophone and computer} - ICMC, Habana, Cuba
 \item[2001] \emph{Saxophone} - Bell Atlantic Jazz Festival, New York. \emph{Michael Formanek
 Northern Exposure}.
 \item[2000] \emph{Saxophone} - Scandinavian tour as well as New York
 performances. \emph{Michael Formanek Northern Exposure}.
 \item[1996-2005] \emph{Saxophone and computer} - Several installations and performances in Scandinavia,
 Germany and Bellarus with Swedish visual artist Stefan Lundgren.
 \item[1997] \emph{Saxophone} - Tour and recordings with Blue Pages, nominated Swedish Jazz
 Group of the Year.
 \item[1996-98] \emph{Saxophone} - Tours and recordings with pianist Richie Beirach.
 \item[1995] \emph{Saxophone} - Swedish tour with David Liebman.
 \item[1992] \emph{Conductor, saxophone} - Montreux International Jazz Festival. \emph{The Orchestra}
\end{enumerate}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "vetenskaplig-meritportfolj"
%%% End: 
