% 
% För att å, ä, ö ska fungera vid pdf output, kör
% >% pdflatex filnamn.tex
%

\documentclass[a4paper,mhm,color,english]{luletter}

\begin{document}

\documentname{Submission}
\diarynumber{}
%\renewcommand{\today}{30 februari 1712}
\name{Henrik Frisk, PhD candidate}
\signature{Henrik Frisk}
\telephone{+46 73 336 83 40}
\phone{2 54 50}
\email{henrik.frisk@mhm.lu.se}
\homepage{http://www.henrikfrisk.com/}

\begin{letter}{Dr Dave Collins\\School of Intermedia \& Performance Arts\\Doncaster College\\High Melton\\Doncaster\\DN5 7SZ\\UK}

\opening{Dear Mr. Collins}
\small

Please find enclosed CD's representing the work of guitarist Stefan
 Östersjö and myself, Henrik Frisk, saxophone, composer and
 improvisor. My biography is attached, and can also be found at
 http://www.henrikfrisk.com and Stefan Östersjö's biography can be found
 in the booklet of his CD.

We would like to suggest a performance as well as a paper presentation,
 the abstract for which we have already e-mailed to you. The performance
 would present my work that I am doing with interactive computer systems
 as well as some of Stefan's work as a performer of works for guitar and
 electronics. One specific piece we would like to propose is the one piece by Love Mangs which we refer to in our paper. This piece was premiered on a Scandinavian tour last year, and is written for guitar, banjo, e-bow and computer (in MAX MSP). Furthermore Stefan and myself have worked together on the issue of interpretation of 'mixed works' hence our performance could form the basis for a workshop on the performance of live electronic music with acoustic instruments and computer.


\closing{Sincerely yours,}

\end{letter}
\end{document}
